## Flow-Control

Sometimes in the course of data analysis we have to make decisions or branch decisions based on what is contained within the data.
The logic of how this done within the programming language is called `flow control`. More generally, `flow control` is an essential aspect of programming that allows you to control the order in which statements and functions are executed.


### For Loop
In R, a loop is a programming construct that allows you to execute a block of code repeatedly. Loops are used when you want to perform a set of instructions repeatedly, such as when you want to iterate over a set of data and perform a particular operation on each element.

There are several types of loops available in R, including the `for` loop, the `while` loop, and the `repeat` loop.

The `for` loop is the most commonly used loop in R. It is used to iterate over a sequence of values, such as a vector or a list, and perform a particular operation on each element of the sequence. The basic syntax of a for loop in R is as follows:
```{r, eval=FALSE}
for (var in sequence) {
  # code to be executed
}

```

Let's look at a very simple `for` loop:
```{r}
# Create a vector of numbers
numbers <- c(1, 2, 3, 4, 5)

# Iterate over the vector using a for loop
for (num in numbers) {
  print(num)
}

```
This code first creates some example data, numbers 1 to 5 increasing by 1, then we go every number and `print` the value. Translating the synatx to english we can say `for` every `num` in `numbers` vector, `print` the `num`. 

### If-Else

When we have data that needs to be treated based on a condition of the data, we have a branching decision. In this case, our flow control is an If-Else statement. In plain english, `if` a condition is met, we do something, `else` we do something else. The brackets between these statements determine what is done.

Let's take the previous example and print if the number is even or odd. You can see already that in how we formulate the programming question we already see the syntax. We say "`if` the number is even..." `if` is our flow control 

```{r}
# Create a vector of numbers
numbers <- c(1, 2, 3, 4, 5)

# Iterate over the vector using a for loop
for (num in numbers) {
  if (num %% 2 == 0) {
    print(paste(num, "is even"))
  } else {
    print(paste(num, "is odd"))
  }
}
```
In this example, we're using the %% operator to determine whether each number in the vector is even or odd. If the number is even, the code inside the if block is executed and the number is printed along with the message "is even". If the number is odd, the code inside the else block is executed and the number is printed along with the message "is odd". `paste is a handy function for combining 