## Exercises {-}

1. Pipeline (eg. `%>%`) a data operation that provides the mean of following vector.
```{r, echo=FALSE}
set.seed(5)
vec_data <- rnorm(10, mean = 10, sd = 3) %>% round(2)
vec_data 
```
```{r, echo=FALSE}
vec_data %>% mean()
```

2.  Employing a pipeline (eg. `%>%`), construct a tibble with columns named `circ` and `diam` which contains the diameters of circles with integer circumferences 1 to 5. Remember PEMDAS.
```{r, echo=FALSE}
library(tidyverse)

tbl_cir <- tibble(
  circ = 1:5
) %>%
  mutate(diam = circ ^ 2 * pi)

tbl_cir
```

3.  Extract all circumferences greater than 50.
```{r, echo=FALSE}
tbl_cir %>% filter(diam > 50)
```

4. Add a column named `circ_type` where you assign the string *odd* or *even* depending on the column `circ`. Attempt to use the `purrr::map` function, along with the `oddeven()` function from the previous chapter, then compute the mean, standard deviation, and coefficient of variation of the diameters for each `circ_type`.
```{r, echo=FALSE}
oddeven <- function(x) { 
  if(x %% 2 == 0) return('even')
  else return('odd')
}

tbl_cir %>% 
  mutate(circ_type = map(circ, oddeven)) %>%
  mutate(circ_type = unlist(circ_type)) %>%
  group_by(circ_type) %>%
  summarise(
    diam_mean = mean(diam),
    diam_sd = sd(diam),
    diam_cv = diam_sd / diam_mean
  )
```