## MSstats

The MSstats package is an R package designed for the analysis of label-free mass spectrometry data. It provides a wide range of statistical tools for the analysis of protein abundance data, including normalization, missing value imputation, quality control, and differential expression analysis. MSstats provides a powerful and flexible way to analyze mass spectrometry data, making it an essential tool for researchers in the field.

|                   |                                                                                                                   |
|----------------|--------------------------------------------------------|
| **Documentation** | [web](https://msstats.org), [PDF](https://bioconductor.org/packages/release/bioc/manuals/MSstats/man/MSstats.pdf) |
| **Code**          | [Bioconductor](https://www.bioconductor.org/packages/release/bioc/html/MSstats.html)                              |
| **Literature**    | *Bioinformatics* 30.17 (2014): 2524-2526                                                                          |

### Installation {-}

To install MSstats, you can use the following code:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("MSstats")
```

### Loading Data {-}

To load your data into MSstats, you can use the following code:

```{r, eval=FALSE}
library(MSstats)
data("QCMAPData")
data <- QCMAPData$data
design <- QCMAPData$design
head(data)
```

### Quality Control {-}

To perform quality control on your data, you can use the following code:

```{r, eval=FALSE}
QCmetrics <- MSstats:::calculateQCmetrics(data, featureLevel = "protein")
MSstats:::plotQCmetrics(QCmetrics, plotType = "boxplot",
                        ylim = c(0, 200000), ylab = "Total Ion Current")
```

### Differential Expression Analysis {-}

To perform differential expression analysis on your data, you can use the following code:

```{r, eval=FALSE}
MSstatsResult <- MSstats::MSstats(data, design,
                                  normalizationMethod = "quantile",
                                  normalization = "median",
                                  msLevel = 1, verbose = TRUE)
summary(MSstatsResult)
```

### Conclusion {-}

Overall, MSstats is a powerful and flexible R package for the analysis of label-free mass spectrometry data. It provides a wide range of statistical tools for the analysis of protein abundance data, making it an essential tool for researchers in the field. With its easy-to-use interface and comprehensive documentation, MSstats is a must-have tool for anyone working with mass spectrometry data.
