## **6** Tidyverse {-}

1. Read in the `bacterial-metabolites_dose-simicillin_tidy.csv` data set.

```{r, eval=FALSE}
url <- "https://raw.githubusercontent.com/jeffsocal/ASMS_R_Basics/main/data/bacterial-metabolites_dose-simicillin_tidy.csv"
download.file(url, destfile = "./data/bacterial-metabolites_dose-simicillin_tidy.csv")
```

```{r, echo=TRUE}
library(tidyverse)
dat <- read_csv("data/bacterial-metabolites_dose-simicillin_tidy.csv")
```


2. How many organisms, metabolites, dose levels, and time points are in the data? How many rows are in the data table? What is the overall study design?

```{r, echo = TRUE}
orgs <- unique(dat$Organism)
n_orgs <- length(unique(orgs))

metabs <- unique(dat$Metabolite)
n_metabs <- length(unique(metabs))

doses <- unique(dat$Dose_mg)
n_doses <- length(unique(doses))

time_pts <- unique(dat$Time_min)
n_time_pts <- length(unique(time_pts))

nrow(dat)
nrow(dat) == n_orgs * n_metabs * n_doses * n_time_pts
```

3. Which metabolite has the highest overall mean abundance?

```{r, echo = TRUE}
dat %>%
  group_by(Metabolite) %>%
  summarize(mean_abundance = mean(Abundance)) %>%
  arrange(desc(mean_abundance))
```

4. Does this metabolite have the highest mean abundance for each organism, or is there differences between organisms?

```{r, echo = TRUE}
dat %>%
  group_by(Organism, Metabolite) %>%
  summarize(mean_abundance = mean(Abundance)) %>%
  arrange(Organism, desc(mean_abundance))
```

5. Is there an overall trend of mean abundance values vs. time point? What about abundance vs. dose?

```{r, echo = TRUE}
dat %>%
  group_by(Time_min) %>%
  summarize(mean_abundance = mean(Abundance))


dat %>%
  group_by(Dose_mg) %>%
  summarize(mean_abundance = mean(Abundance))
```

6. Using the example code at the beginning of this Chapter (using the Dever climate data), compute a linear fit of log10 abundance vs. time point for each metabolite and plot the results.

```{r, echo = TRUE}
dat <- dat %>%
  mutate(log10_abundance = log10(Abundance))

lm_func <- function(data) {
  lm(log10_abundance ~ Time_min, data = data)
}


dat_lm <- dat %>%
  # dplyr
  group_by(Metabolite) %>%
  # tidyr
  nest() %>%
  # dplyr, purrr: apply the function to each nested data frame
  mutate(model = map(data, lm_func)) %>%
  # dplyr, broom, purrr: extract the coefficients from each model
  mutate(tidy = map(model, broom::tidy)) %>%
  # tidyr
  unnest(tidy) %>%
  ungroup() %>%
  # dplyr, stringr: clean-up the terms
  mutate(term = term %>% str_replace_all("\\(|\\)", "")) %>%
  # dplyr: retain only specific columns
  select(Metabolite, term, estimate) %>%
  # tidyr: convert from a long table to a wide table
  pivot_wider(names_from = 'term', values_from = 'estimate') %>%
  # dplyr: create a new column with the model slope (better name), just copy Time_min
  mutate(model_slope = Time_min)


#ggplot2
ggplot(dat, aes(Time_min, log10_abundance)) + 
  # represent the data as points
  geom_point() +
  # use the linear model data to plot regression lines
  geom_abline(data = dat_lm,
              aes(slope = model_slope, intercept = Intercept)) +
  # plot each year separately 
  facet_wrap(~Metabolite)

```

